Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-04 paper/main.tex
Updates

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on their domain knowledge. This preference arises due to several
practical issues with automated algorithms, such as they make obvious errors
and out Markov Equivalence Classes (MECs). To address these challenges, we
propose an iterative method to assist researchers in manually constructing or
improving an existing DAG by combining implied Conditional Independence (CI)
testing with expert knowledge. Our proposed method utilizes a novel measure of
partial association for mixed data to rank violations of implied CIs in a given
DAG, enabling users to prioritize modifications while utilizing their expertise
when making these modifications. Empirical results show that this manual model
construction approach achieves performance comparable to automated algorithms
even if the user is able to decide the orientation of an edge correctly in one
out of three cases. Additionally, we demonstrate that, in the absence of an
expert, a Large Language Model (LLM) can potentially determine edge orientations.

>>>>

Causal discovery has received significant attention in the Directed Acyclic
Graphs (DAGs) literature, leading to the development of numerous automated
algorithms for learning DAGs from data. However, their adoption in applied
domains remain limited, as researchers often prefer to construct DAGs manually
based on their domain knowledge. This preference arises due to several
practical issues with automated algorithms, such as they make obvious errors
and out Markov Equivalence Classes (MECs). To address these challenges, we
propose an iterative method to assist researchers in manually constructing or
improving an existing DAG by combining implied Conditional Independence (CI)
testing with expert knowledge. Our proposed method utilizes a novel measure of
partial association for mixed data to rank violations of implied CIs in a given
DAG, enabling users to prioritize modifications while utilizing their domain
knowledge to decide the orientation of new edges. Empirical results show that
this manual model construction approach achieves performance comparable to
automated algorithms even if the user is able to decide the edge orientation
correctly in one out of three cases. Additionally, we demonstrate that, in the
absence of an expert, a Large Language Model (LLM) can be potentially determine
edge orientations. We provide the implementation of our approach in both a web 
tool at: <redacted for review> and a Python package <redacted for review>.
